---
title: "Post-AKI Survival Analysis"
author: "Rui Feng"
date: "7/25/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)
```

# Overview

This markdown prepares the post-AKI cohorts, describes CKD-free survival and overall survival, performs Elastic Net variable selection, fits competing-risk regression models for CKD and mortality, evaluates model performance in MIMIC-IV, externally validates the models in MIMIC-III, and produces time-dependent AUC and Brier-score plots with 95% confidence limits.

# Packages

```{r packages}
library(arrow)          # Read Parquet files
library(survival)       # Surv() and Kaplan-Meier estimation
library(glmnet)         # Penalized Cox regression
library(glmnetUtils)    # Formula interface and cva.glmnet()
library(riskRegression) # Fine-Gray models, Score(), AUC, and Brier scores
library(prodlim)        # Hist() competing-risk outcome
library(dplyr)          # Data manipulation
library(tidyr)          # Data reshaping
library(purrr)          # Functional programming helpers
library(data.table)     # Efficient data handling used by riskRegression
library(ggplot2)        # Graphics
library(patchwork)      # Multi-panel figures
library(scales)         # Percentage axis labels
```

# Analysis settings

```{r analysis-settings}
set.seed(20260627)

analysis_file <- "features_ckd_imputed.parquet"
follow_up_times <- 90 * seq_len(40)
sex_colors <- c("Female" = "red", "Male" = "blue")
```

# Import and prepare the post-AKI cohort
The event code is defined as 0 for censoring, 1 for CKD, and 2 for death. Post-CKD events are combined with CKD. The original code removed five reference-category columns by position; those positions are retained here for reproducibility, although removing variables by name is preferable when names are available.

```{r import-preprocess}
mimic_ckd <- read_parquet(analysis_file) %>%
  as.data.frame()

# Review event distributions before recoding.
with(
  mimic_ckd,
  table(event_type, version, time_days >= 90, useNA = "ifany")
)

# Combine PostCKD with CKD and create a competing-risk event code.
mimic_ckd <- mimic_ckd %>%
  mutate(
    is_ckd = if_else(is_postckd %in% TRUE, 1, is_ckd),
    event_type = if_else(event_type == "PostCKD", "CKD", event_type),
    event_code = case_when(
      event_type == "CKD"   ~ 1L,
      event_type == "Death" ~ 2L,
      TRUE                  ~ 0L
    )
  )

with(mimic_ckd, table(is_ckd, event_code, useNA = "ifany"))

# Remove reference-category columns used in the original analysis.
reference_columns <- c(26, 30, 37, 39, 46)
if (max(reference_columns) <= ncol(mimic_ckd)) {
  mimic_ckd <- mimic_ckd[, -reference_columns, drop = FALSE]
}

# Separate the development and external-validation cohorts.
mimic3_ckd <- mimic_ckd %>% filter(version == "mimic3")
mimic4_ckd <- mimic_ckd %>% filter(version == "mimic4")

cat("MIMIC-III post-AKI cohort:", nrow(mimic3_ckd), "patients\n")
cat("MIMIC-IV post-AKI cohort:", nrow(mimic4_ckd), "patients\n")
```

# Kaplan-Meier curves by sex

K-M curves compare CKD-free survival and overall survival by sex across MIMIC-III and MIMIC-IV. For CKD, death is treated as censoring in this descriptive Kaplan-Meier analysis; the formal CKD models below account for death as a competing event.

```{r kaplan-meier, fig.width=11, fig.height=5}
km_ckd3 <- survfit(
  Surv(time_days / 365.25, event_type == "CKD") ~ gender,
  data = mimic3_ckd
)
km_ckd4 <- survfit(
  Surv(time_days / 365.25, event_type == "CKD") ~ gender,
  data = mimic4_ckd
)

km_death3 <- survfit(
  Surv(time_days / 365.25, is_death) ~ gender,
  data = mimic3_ckd
)
km_death4 <- survfit(
  Surv(time_days / 365.25, is_death) ~ gender,
  data = mimic4_ckd
)

x_limits <- c(90 / 365.25, 10)
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 2, 1))

plot(
  km_ckd3,
  col = unname(sex_colors), lty = 1, lwd = 2,
  conf.int = FALSE, mark.time = FALSE,
  xlab = "Years since AKI admission",
  ylab = "Probability of remaining CKD-free",
  main = "CKD after AKI",
  xlim = x_limits
)
lines(
  km_ckd4,
  col = unname(sex_colors), lty = 3, lwd = 2,
  conf.int = FALSE, mark.time = FALSE
)
legend(
  "bottomleft",
  legend = c(
    "MIMIC-III Female", "MIMIC-III Male",
    "MIMIC-IV Female", "MIMIC-IV Male"
  ),
  col = c("red", "blue", "red", "blue"),
  lty = c(1, 1, 3, 3), lwd = 2,
  bty = "n", cex = 0.85
)

plot(
  km_death3,
  col = unname(sex_colors), lty = 1, lwd = 2,
  conf.int = FALSE, mark.time = FALSE,
  xlab = "Years since AKI admission",
  ylab = "Probability of survival",
  main = "Death after AKI",
  xlim = x_limits,
  ylim = c(0, 1)
)
lines(
  km_death4,
  col = unname(sex_colors), lty = 3, lwd = 2,
  conf.int = FALSE, mark.time = FALSE
)
legend(
  "topright",
  legend = c(
    "MIMIC-III Female", "MIMIC-III Male",
    "MIMIC-IV Female", "MIMIC-IV Male"
  ),
  col = c("red", "blue", "red", "blue"),
  lty = c(1, 1, 3, 3), lwd = 2,
  bty = "n", cex = 0.85
)

par(mfrow = c(1, 1))
```

# Construct candidate predictors and sex interactions

The original analysis selected baseline variables and laboratory summary. It then created interactions between sex and all candidate predictors except the sex indicator itself.

```{r feature-construction}
prepare_model_data <- function(data) {
  n_markers <- 11
  n_repeats <- 9

  laboratory_columns <-
    44 + rep(seq_len(n_markers) - 1, each = 1) * n_repeats +
    rep(3, n_markers)

  selected_columns <- c(18:42, laboratory_columns)

  if (max(selected_columns) > ncol(data)) {
    stop("One or more selected column positions exceed the number of columns in the dataset.")
  }

  predictors <- data[, selected_columns, drop = FALSE]
  predictors <- predictors[, -(19:20), drop = FALSE]

  # The 12th selected predictor is the male-sex indicator in the source data.
  sex_position <- 12
  if (sex_position > ncol(predictors)) {
    stop("The expected sex indicator was not found at position 12.")
  }

  interaction_terms <- predictors[, -sex_position, drop = FALSE]
  interaction_terms <- sweep(
    as.matrix(interaction_terms),
    MARGIN = 1,
    STATS = predictors[[sex_position]],
    FUN = "*"
  ) %>%
    as.data.frame()

  names(interaction_terms) <- paste0(names(interaction_terms), "M")

  bind_cols(
    data %>% select(time_days, event_code),
    predictors,
    interaction_terms
  )
}

mimic4_model <- prepare_model_data(mimic4_ckd)
mimic3_model <- prepare_model_data(mimic3_ckd)
```

# Elastic Net variable selection

Elastic Net Cox models are used to screen candidate main effects and sex interactions separately for CKD and mortality. The alpha parameter is searched using cross-validated Harrell's C statistic. The final alpha and lambda values match those selected in the original analysis.

## CKD variable selection

```{r elastic-net-ckd}
alpha_grid_ckd <- seq(0, 0.1, length.out = 11)

cv_ckd <- cva.glmnet(
  Surv(time_days, event_code == 1) ~ .,
  data = mimic4_model,
  family = "cox",
  alpha = alpha_grid_ckd,
  nfolds = 10,
  type.measure = "C"
)

plot(cv_ckd)

elastic_ckd <- glmnet(
  Surv(time_days, event_code == 1) ~ .,
  data = mimic4_model,
  family = "cox",
  alpha = 0.02,
  lambda = exp(-1.5)
)

ckd_coefficients <- as.matrix(coef(elastic_ckd))
selected_ckd <- rownames(ckd_coefficients)[ckd_coefficients[, 1] != 0]
selected_ckd
```

## Mortality variable selection

```{r elastic-net-death}
alpha_grid_death <- seq(0, 0.1, length.out = 21)

cv_death <- cva.glmnet(
  Surv(time_days, event_code == 2) ~ .,
  data = mimic4_model,
  family = "cox",
  alpha = alpha_grid_death,
  nfolds = 20,
  type.measure = "C"
)

plot(cv_death)

elastic_death <- glmnet(
  Surv(time_days, event_code == 2) ~ .,
  data = mimic4_model,
  family = "cox",
  alpha = 0.02,
  lambda = exp(-1)
)

death_coefficients <- as.matrix(coef(elastic_death))
selected_death <- rownames(death_coefficients)[death_coefficients[, 1] != 0]
selected_death
```

# Competing-risk models for CKD after AKI

Fine-Gray regression estimates subdistribution hazard ratios for CKD while treating death as a competing event. The baseline model includes demographic and comorbidity predictors. The extended model adds laboratory measures and sex-predictor interactions.

```{r fine-gray-ckd}
ckd_baseline <- FGR(
  Hist(time_days, event_code) ~
    age + cmb_htn + cmb_dm + cmb_cad + cmb_hf + cmb_liver +
    dem_sex_M + dem_mar_WIDOWED,
  data = mimic4_model,
  cause = 1
)

ckd_extended <- FGR(
  Hist(time_days, event_code) ~
    age + cmb_htn + cmb_dm + cmb_cad + cmb_hf + cmb_liver +
    dem_sex_M + dem_mar_WIDOWED +
    labs90d_bun_mean_90d + labs90d_lactate_mean_90d +
    labs90d_ph_mean_90d + dem_race_BLACKM +
    labs90d_bun_mean_90dM + labs90d_creatinine_mean_90dM,
  data = mimic4_model,
  cause = 1
)

summary(ckd_baseline)
summary(ckd_extended)
```

# Evaluate CKD models

MIMIC-IV is used for model development and internal cross-validation. MIMIC-III is used as an external validation cohort without refitting the models.

```{r evaluate-ckd}
set.seed(20260627)

score_ckd_dev <- Score(
  object = list(Base = ckd_baseline, Full = ckd_extended),
  formula = Hist(time_days, event_code) ~ 1,
  data = mimic4_model,
  cause = 1,
  times = follow_up_times,
  metrics = c("AUC", "Brier"),
  conf.int = TRUE,
  contrasts = TRUE,
  split.method = "cv5",
  B = 50,
  cens.model = "km",
  null.model = TRUE
)

score_ckd_val <- Score(
  object = list(Base = ckd_baseline, Full = ckd_extended),
  formula = Hist(time_days, event_code) ~ 1,
  data = mimic3_model,
  cause = 1,
  times = follow_up_times,
  metrics = c("AUC", "Brier"),
  conf.int = TRUE,
  contrasts = TRUE,
  split.method = "none",
  cens.model = "km",
  null.model = TRUE
)

plotAUC(score_ckd_val)
plotBrier(score_ckd_val)
```

# Competing-risk models for death after AKI

For the mortality analysis, death is the event of interest and CKD is treated as the competing event. The extended model adds laboratory variables and a heart-failure-by-sex interaction.

```{r fine-gray-death}
death_baseline <- FGR(
  Hist(time_days, event_code) ~
    age + cmb_htn + cmb_dm + cmb_hf + cmb_copd + cmb_cancer +
    cmb_infection + cmb_liver + cmb_cerebrovasc + cmb_obesity +
    dem_sex_M + dem_race_BLACK + dem_race_OTHER + dem_mar_UNK,
  data = mimic4_model,
  cause = 2
)

death_extended <- FGR(
  Hist(time_days, event_code) ~
    age + cmb_htn + cmb_dm + cmb_hf + cmb_copd + cmb_cancer +
    cmb_infection + cmb_liver + cmb_cerebrovasc + cmb_obesity +
    dem_sex_M + dem_race_BLACK + dem_race_OTHER + dem_mar_UNK +
    labs90d_wbc_mean_90d + labs90d_hgb_mean_90d +
    labs90d_plt_mean_90d + labs90d_potassium_mean_90d +
    labs90d_creatinine_mean_90d + labs90d_bicarbonate_mean_90d +
    labs90d_lactate_mean_90d + cmb_hfM,
  data = mimic4_model,
  cause = 2
)

summary(death_baseline)
summary(death_extended)
```

# Evaluate mortality models

```{r evaluate-death}
set.seed(20260627)

score_death_dev <- Score(
  object = list(Base = death_baseline, Full = death_extended),
  formula = Hist(time_days, event_code) ~ 1,
  data = mimic4_model,
  cause = 2,
  times = follow_up_times,
  metrics = c("AUC", "Brier"),
  conf.int = TRUE,
  contrasts = TRUE,
  split.method = "cv5",
  B = 50,
  cens.model = "km",
  null.model = TRUE
)

score_death_val <- Score(
  object = list(Base = death_baseline, Full = death_extended),
  formula = Hist(time_days, event_code) ~ 1,
  data = mimic3_model,
  cause = 2,
  times = follow_up_times,
  metrics = c("AUC", "Brier"),
  conf.int = TRUE,
  contrasts = TRUE,
  split.method = "none",
  cens.model = "km",
  null.model = TRUE
)

plotAUC(score_death_val)
plotBrier(score_death_val)
```

# Save fitted model-performance objects

```{r save-models}
saveRDS(score_ckd_dev,   "score_ckd_mimic4.rds")
saveRDS(score_ckd_val,   "score_ckd_mimic3.rds")
saveRDS(score_death_dev, "score_death_mimic4.rds")
saveRDS(score_death_val, "score_death_mimic3.rds")
```

# Prepare performance estimates and confidence limits

The helper functions accommodate minor differences in confidence-limit column names across versions of `riskRegression`.

```{r tidy-score-functions}
find_column <- function(data, candidates, label) {
  matched <- intersect(candidates, names(data))
  if (length(matched) == 0) {
    stop(
      label, " column was not found. Available columns: ",
      paste(names(data), collapse = ", ")
    )
  }
  matched[[1]]
}

standardize_model_label <- function(x) {
  case_when(
    grepl("^Null", x, ignore.case = TRUE) ~ "Null",
    grepl("^Base", x, ignore.case = TRUE) ~ "Base",
    grepl("^Full", x, ignore.case = TRUE) ~ "Full",
    TRUE ~ as.character(x)
  )
}

tidy_score_metric <- function(score_object, dataset_name, metric) {
  metric <- match.arg(metric, c("AUC", "Brier"))
  score_table <- as.data.frame(score_object[[metric]]$score)

  time_column <- find_column(score_table, c("times", "time"), "Time")
  lower_column <- find_column(
    score_table,
    c("lower", "lower.95", "lower95", "ci.lower"),
    "Lower confidence limit"
  )
  upper_column <- find_column(
    score_table,
    c("upper", "upper.95", "upper95", "ci.upper"),
    "Upper confidence limit"
  )

  value_column <- if (metric == "AUC") "AUC" else "Brier"

  score_table %>%
    transmute(
      Dataset = dataset_name,
      Model = standardize_model_label(as.character(model)),
      times = .data[[time_column]],
      estimate = as.numeric(.data[[value_column]]),
      lower = pmax(as.numeric(.data[[lower_column]]), 0),
      upper = pmin(as.numeric(.data[[upper_column]]), 1),
      time_years = times / 365.25
    ) %>%
    filter(
      is.finite(estimate),
      is.finite(lower),
      is.finite(upper)
    )
}

combine_performance <- function(development_score, validation_score, metric) {
  bind_rows(
    tidy_score_metric(development_score, "MIMIC-IV (D)", metric),
    tidy_score_metric(validation_score, "MIMIC-III (V)", metric)
  ) %>%
    mutate(
      Dataset = factor(
        Dataset,
        levels = c("MIMIC-IV (D)", "MIMIC-III (V)")
      )
    )
}
```

```{r build-performance-data}
ckd_auc <- combine_performance(score_ckd_dev, score_ckd_val, "AUC") %>%
  filter(Model %in% c("Base", "Full")) %>%
  mutate(Model = factor(Model, levels = c("Base", "Full")))

ckd_brier <- combine_performance(score_ckd_dev, score_ckd_val, "Brier") %>%
  filter(Model %in% c("Null", "Base", "Full")) %>%
  mutate(Model = factor(Model, levels = c("Null", "Base", "Full")))

death_auc <- combine_performance(score_death_dev, score_death_val, "AUC") %>%
  filter(Model %in% c("Base", "Full")) %>%
  mutate(Model = factor(Model, levels = c("Base", "Full")))

death_brier <- combine_performance(score_death_dev, score_death_val, "Brier") %>%
  filter(Model %in% c("Null", "Base", "Full")) %>%
  mutate(Model = factor(Model, levels = c("Null", "Base", "Full")))

max_years <- ceiling(
  max(
    c(
      ckd_auc$time_years,
      ckd_brier$time_years,
      death_auc$time_years,
      death_brier$time_years
    ),
    na.rm = TRUE
  )
)
year_breaks <- seq(0, max_years, by = 1)
```

# Plot time-dependent AUC and Brier scores

Thin outer lines represent the lower and upper 95% confidence limits. Model type is encoded by color and dataset by line type.

```{r performance-plot-function}
auc_palette <- c("Base" = "#1b9e77", "Full" = "purple")
brier_palette <- c(
  "Null" = "grey60",
  "Base" = "#1b9e77",
  "Full" = "purple"
)
dataset_linetypes <- c(
  "MIMIC-IV (D)" = "solid",
  "MIMIC-III (V)" = "dashed"
)

plot_performance <- function(
    data,
    y_label,
    palette,
    y_limits = NULL,
    y_breaks = waiver(),
    show_legend = TRUE,
    percent_axis = FALSE) {

  plot_object <- ggplot(
    data,
    aes(
      x = time_years,
      y = estimate,
      color = Model,
      linetype = Dataset,
      group = interaction(Model, Dataset)
    )
  ) +
    geom_line(aes(y = lower), linewidth = 0.35) +
    geom_line(aes(y = upper), linewidth = 0.35) +
    geom_line(linewidth = 1) +
    scale_color_manual(values = palette, drop = FALSE) +
    scale_linetype_manual(values = dataset_linetypes) +
    scale_x_continuous(
      breaks = year_breaks,
      labels = year_breaks,
      limits = c(0, max_years)
    ) +
    labs(x = "Years", y = y_label, title = NULL) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = if (show_legend) "bottom" else "none",
      legend.title = element_blank(),
      panel.grid.minor = element_blank()
    )

  if (percent_axis) {
    plot_object <- plot_object +
      scale_y_continuous(
        limits = y_limits,
        breaks = y_breaks,
        labels = percent_format(accuracy = 1),
        expand = expansion(mult = c(0, 0.02))
      )
  } else {
    plot_object <- plot_object +
      scale_y_continuous(
        limits = y_limits,
        breaks = y_breaks,
        expand = expansion(mult = c(0, 0.02))
      )
  }

  plot_object
}
```

```{r four-panel-performance-figure, fig.width=12, fig.height=9}
ckd_auc_plot <- plot_performance(
  ckd_auc,
  y_label = "Time-dependent AUC",
  palette = auc_palette,
  y_limits = c(0.3, 1.0),
  y_breaks = seq(0.3, 1.0, by = 0.1),
  show_legend = TRUE,
  percent_axis = TRUE
)

ckd_brier_plot <- plot_performance(
  ckd_brier,
  y_label = "Brier score",
  palette = brier_palette,
  y_limits = c(0, max(ckd_brier$upper, na.rm = TRUE) * 1.05),
  show_legend = FALSE
)

death_auc_plot <- plot_performance(
  death_auc,
  y_label = "Time-dependent AUC",
  palette = auc_palette,
  y_limits = c(0.6, 0.9),
  y_breaks = seq(0.6, 0.9, by = 0.1),
  show_legend = TRUE,
  percent_axis = TRUE
)

death_brier_plot <- plot_performance(
  death_brier,
  y_label = "Brier score",
  palette = brier_palette,
  y_limits = c(0, max(death_brier$upper, na.rm = TRUE) * 1.05),
  show_legend = FALSE
)

(
  (ckd_auc_plot | ckd_brier_plot) /
  (death_auc_plot | death_brier_plot)
) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 13))
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
